Versioned data: why it’s needed and how it can be achieved (easily and cheaply)

### 2017-08-03 ms.tex
Figure illustrating technologies and users
For #11
Minor edits and more references

Data is now a first-class scientific product. Funding bodies, publishers, and scientific social norms are increasingly both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses. This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators [Cite some Open Science papers, both pro and con]. One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous. On one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}). On the other end are repositories for storing a wide variety of static, stand-alone datasets (\texttt{Dryad}, \texttt{Figshare}, and less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}). We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.

In this article we i) introduce the concept of Versioned Data Delivery (VDD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be leveraged to help researchers maintain, distribute, and access small to medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a VDD system. While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. Furthermore, our backgrounds in the life sciences (and ecology and evolution, in particular); however, we argue that the issues we identify and the solutions we put forward are likely to be relevant to a wide variety of domains, across the natural and social sciences.

\subsection{Why versioning of data is needed}

@@

\subsection{Why versioning of data has been difficult}

While the benefits of versioned data are fairly clear, determining how best to implement it has been anything but. The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
or under its own DOI, is simply not designed to provide streams of versioned
data (Table \ref{tab:sql_v_versioneddata}). A particular issue with the static
model is that data releases are mostly triggered by publication of academic
papers. But for a living dataset, there may be multiple releases made in
between papers. This is particularly the case with the discovery of small
errors. For any medium-sized dataset that is put to reuse, errors will
inevitably be found, and there needs to be some easy  way to correct them and
distribute the updated dataset.

@@

When considering the right tool to store and deliver versioned data, the size
of the dataset is another essential consideration. ``Big data'' brings with it
specific needs and so requires a specific set of hardware and software tools.
Examples of big data include genetic datasets, global datasets of species
observations, and remote sensing data. This size of this data is bigger than
that which individual research groups can maintain and often requires a
specialized governmental or non-governmental institution with full-time staff
to curate and maintain those research resources. Here we are particularly
interested in smaller and more specialized problems (e.g., cases where a single 
research group contributes most of the data). In most of these cases, the data
collected will not be ``big'' but rather small to medium sized. Such data
support research projects, but are not in most cases general enough to
establish infra-structure for. But crucially the data themselves will change
over time. And currently there is no infrastructure for organizing and
distributing these types of ``living datasets''.

@@

In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software development and applies them to the challenge of maintaining and distributing versioned data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local  computers. With a VDD system, data developers maintain a core set of files, which produce an organised dataset that can be similarly be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, VDD uses well-established and tools, which ensures high-level performance and stability.

@@

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.  Essentially version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small to medium sized datasets \citep{Ram-2013, Perkel-2016}. Git is attractive for data management because it tracks all changes in monitored files. This change history is visible to anyone interacting with the repository. It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.

Because \texttt{git} was built from the ground up for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.  There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future (REF).  This future system would be faster computationally, smaller in file size, and thus allow for larger datasets to be versioned efficiently.  That said, the workflow we describe here currently works well with \texttt{git} and should work more smoothly with its future replacement.

@@

To fully realize the benefits of versioned data, the versioning system needs to be communicated to the user.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is some benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise[] maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.

@@

At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements})).  GitHub has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those "releases" can be downloaded directly by users via point-and-click.  Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.  Releases can also be linked to code-based users via an R package which we describe in the following section.

@@

For many use cases, a point-and-click workflow may create barriers for reproducibility. Ideally, users should be able to also able access the all versions of the database programmatically. To this end, we offer a novel solution. We have written a R package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the GitHub API to pull down versions of the dataset; and b) constructs the shell of a second, database specific R package, which can be used to access the dataset. Using this system, a researcher can create and distribute a R package that facilitates access to her data with (very) minimal computational skills.

@@

%Topics to consider:
%
%\begin{itemize}
%
%\item Addresses important need, leveraging (use) open source platforms. Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
%\item Versioning and open access essential for reproducibility
%\item Versioning should enable better benchmarking of models, also because tests can be automated.
%\item We have focussed on particular set of tools, but could do similar things with other platforms or languages.
%....
%\end{itemize}

@@

\begin{figure}[!hb]
\centering
%\includegraphics[width=\linewidth]{figure.pdf}
\caption{Overview of the technologies involved in distributing Lightweight versioned data.}
\label{fig:technology_stack}
\end{figure}
\bibliography{refs}

>>>>

Data is now a first-class scientific product, as is evidenced by the advent of this journal \cite{Editorial-2014}. Increasingly, funding bodies, publishers, and scientific social norms are both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013}. One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous. At one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}). At the other end are repositories for storing a wide variety of static, stand-alone datasets (eg. \texttt{Dryad}, \texttt{Figshare}, and \texttt{Zenodo}; or less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}). We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.

In this article we i) introduce the concept of Versioned Data Delivery (VDD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a VDD system. While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. While the examples cited here are in ecology and evolution; the issues identified and the solutions put forward are relevant to nearly all data-oriented scientific domains.

\subsection{Why versioning of data is needed and difficult}

@@

While the benefits of versioned data are fairly clear, determining how to implement it has been contentious . The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
or under its own DOI, is simply not designed to provide streams of versioned
data (Table \ref{tab:sql_v_versioneddata}). A particular issue with the static
model is that data releases are mostly triggered by publication of academic
papers. But for a living dataset, there may be multiple releases made in
between papers. This is particularly the case with the discovery of small
errors. For any medium-sized dataset that is put to reuse, errors will
inevitably be found, and there needs to be some easy  way to correct them and
distribute the updated dataset.

@@

When considering the right tool to store and deliver versioned data, the size
of the dataset is another essential consideration. ``Big data'' brings with it
specific needs and so requires a specific set of hardware and software tools.
Examples of big data include genetic datasets, global datasets of species
observations, and remote sensing data. This size of this data is bigger than
that which individual research groups can maintain and often requires a
specialized governmental or non-governmental institution with full-time staff
to curate and maintain those research resources. Here we are particularly
interested in smaller and more specialized problems (e.g., cases where a single 
research group contributes most of the data). In most of these cases, the data
collected will not be ``big'' but rather small-to-medium sized. Such data
support research projects, but are not in most cases general enough to
establish bespoke infrastructure for. But crucially the data themselves will change
over time. And currently there is no infrastructure for organizing and
distributing these types of ``living datasets''.

@@

In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local  computers. With a VDD system, data developers maintain a core set of files, which produce an organised dataset that can be similarly be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, VDD uses well-established tools, ensuring high-level performance and stability.

@@

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.  Essentially version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}. Git is attractive for data management because it tracks all changes in monitored files. This history is visible to anyone interacting with the repository. It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.  There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future (REF).  This future system would be faster computationally, smaller in file size, and thus allow for larger datasets to be versioned efficiently.  That said, the workflow we describe here currently works well with \texttt{git} and should work more smoothly with its future replacement.

@@

To fully realize the benefits of a versioned data, the VDD system needs to be communicated to the user.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.

@@

At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  GitHub has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed via the Github API. Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.  Releases can also be linked to code-based users via an R package which we describe in the following section.

@@

A point-and-click workflow may create barriers for reproducibility \cite{Wilson-2014,Lowndes-2017}. Ideally, users should be able to also able access the all versions of the database programmatically. To this end, we offer a novel implementation of a VDD system. We have written a R package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the GitHub API to pull down versions of the dataset; and b) constructs the shell of a second, database specific R package, which can be used to access versions of the dataset. Using this system, a researcher can create and distribute a R package that facilitates access to her data with (very) minimal computational skills.

@@

%Topics to consider:
%
%\begin{itemize}
%
%\item Addresses important need, using open source platforms. Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
%\item Versioning and open access essential for reproducibility
%\item Versioning should enable better benchmarking of models, also because tests can be automated.
%\item We have focussed on particular set of tools, but could do similar things with other platforms or languages.
%....
%\end{itemize}

@@

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{figures/Figure2.pdf}
\caption{Overview of the technologies involved in distributing versioned data via datastorr.}
\label{fig:technology_stack}
\end{figure}
\bibliography{refs}
